\textbf{Advection versus diffusion in intraventricular transport}

### 2025-03-25 paper/main.tex
edits after read-through

\author[1,*]{Halvor Herlyng}
\author[1]{Ada J.~Ellingsrud}
\author[1]{Miroslav Kuchta} 
\author[2]{Inyoung Jeong}
\author[1,3]{Marie E.~Rognes}
\author[2, *]{Nathalie Jurisch-Yaksi}
\affil[1]{Department of Numerical Analysis and Scientific Computing, Simula Research Laboratory, Oslo, Norway} % Kristian Augusts gate 23, 0164
\affil[2]{Department of Clinical and Molecular Medicine, Norwegian University of Science and Technology, Trondheim, Norway} % Erling Skjalgsons gate 1, 7491
\affil[3]{K.~G.~Jebsen Center for Brain Fluid Research, Oslo, Norway}
\affil[*]{\href{mailto:hherlyng@simula.no}{hherlyng@simula.no}, \href{mailto:nathalie.jurisch-yaksi@ntnu.no}{nathalie.jurisch-yaksi@ntnu.no}}
\date{\vspace{-2em}}

% Text: Please use section headings and subheadings as specified below. For communications, all section headings apart from Experimental Section should be removed
% Please make the first reference to a display item bold: \textbf{Figure 1}
% Do not abbreviate Figure, Equation, etc.; display items are always singular, i.e., Figure 1 and 2.
% Equations are always singular, i.e., Equation 1 and 2, and should be inserted using the {equation} environment, not as graphics
% Please do not use footnotes in the text, additional information can be added to the Reference list.

%\section{First Section}

%\subsection{First Subsection}

%\subsubsection{First Sub Subsection}

%\threesubsection{First lowest-level subsection}

@@

% Abstract should be written in the present tense and impersonal style (i.e., avoid we), and be at most 200 words long
\begin{abstract}

    Cerebrospinal fluid (CSF) is integral to brain function, as the fluid
    provides mechanical support for the brain and the flow of CSF helps
    distribute nutrients, neurotransmitters and metabolites throughout
    the central nervous system. CSF flow is partly driven
    by beating of motile cilia, hair-like structures that line the ependymal cell layer
    of the walls of the brain ventricles. Despite the physiological importance of CSF flow,
    the underlying mechanisms of cilia-mediated flow and transport
    in the brain ventricles remain yet to be comprehensively resolved.
    In this study, we analyze and evaluate the role of motile cilia
    in intraventricular CSF flow and transport. We use the geometry
    of the zebrafish larval brain ventricles, for which we have detailed knowledge of cilia properties and CSF motion velocities
    and frequencies, to develop finite element methods that model flow and transport.
    The computational model is validated by in vivo experiments that monitor transport
    of a photoconvertible protein secreted in the brain ventricles.
    Our results show that while cilia contribute to advection of large particles,
diffusion plays a significant role in the transport of small molecules such as proteins.
We also found that cilia location and the geometry of the ventricular system impact solute distribution.
Altogether, our work presents a computational framework that can be applied to other
ventricular systems, together with new concepts of how molecules are transported within the brain and its ventricles.

@@

% I2: What do we know about the problem and the topic?
The flow of CSF in the ventricular system is driven by multiple
physiological processes, including via the beating motion of motile
cilia, CSF secretion by the choroid plexus, and pulsatile pressure gradients generated by the
cardiac cycle, the respiration, neural activity, and bodily
movement~\cite{Brodal2016, del2010ependymal, Ringers2020Role, Vinje2019RespiratoryMeasurements,
Kurtcuoglu2005ComputationalSystem, Olstad2019CiliaryDevelopment,
macaulay2022cerebrospinal, mestre2018flow}.
Cilia are hair-like structures found in a vast diversity of
biological organisms and in several human organs,
of which the brain is one of them~\cite{Fulford1986Muco-ciliaryLung,
Pacherres2022CiliaryProduction, Jahn1972LocomotionProtozoa,
Brennen1977FluidFlagella, Ringers2020Role, spassky2013motile,
DGama2025MotileBrain, mitchell2007evolution,
Tsukita2012CoordinatedFeet, BLAKE1974MechanicsMotion,
Reiten2017Motile-Cilia-MediatedComputations, Thouvenin2020OriginCanal}. 
The ventricular surface of the brain, known as the ependyma, is lined by
motile cilia that beat in a coordinated pattern at frequencies of
10--40 Hz to mediate fluid flow~\cite{mitchell2007evolution,
Ringers2020Role, spassky2013motile, Ringers2023NovelEpithelia, roth2025structure}. 
Given their microscopic size, ranging from
5 to 15 microns, cilia contribute primarily to flow near the
ependyma in the human brain~\cite{Siyahhan2014FlowVentricles, Olstad2019CiliaryDevelopment,
Ringers2020Role, Faubel2016Cilia-basedVentricles}. In contrast, cardiac,
respiratory and delta waves at lower frequencies induce pulsatile flow
at the larger scales in the brain ventricular system, subarachnoid space, and
subarachnoid perivascular
spaces~\cite{Vinje2019RespiratoryMeasurements, eide2024functional,
Causemann2025, fultz2019coupled, mestre2018flow, daversin2020mechanisms}.

@@

% I4: How do we address the problem and what do we find out?
In this study, we combined computational and experimental approaches to investigate
the role of motile cilia in intraventricular CSF flow and transport.
We model the flow and transport by partial differential equations and solve
them numerically with finite element methods to compute flow patterns and
solute concentration dynamics. A geometry representing the zebrafish larval brain ventricles
was used, for which we have detailed knowledge of cilia properties and CSF motion velocities
and frequencies~\cite{Olstad2019CiliaryDevelopment}. 
Notably, this early developmental stage allows us to simplify our model
by omitting CSF secretion, which starts at later stages~\cite{Jeong2024TheZebrafish},
while still being well-conserved with mammals~\cite{DGama2025MotileBrain,
Olstad2019CiliaryDevelopment, Ringers2020Role, DGama2021Diversity, jurisch2020radial}. 
The computational model is validated with experimental data by monitoring in vivo
the dynamics of a photoconvertible fluorescent protein Dendra2
expressed within the brain ventricles.
Our results show that diffusion plays a major role in the transport of small solutes,
whereas cilia majorly contribute to the movement of larger particles.
We also found that the location of cilia, as well as changes to the ventricular geometry,
largely impact solute distribution. In summary, our work presents a computational
framework that can be applied to other
ventricular systems in the future, together with new concepts
of how cilia distribute molecules and
particles within the brain ventricles.

@@

In vivo particle tracking reveals characteristic CSF flow patterns in
zebrafish larval brain ventricles
(\Crefpanels{fig:fig1}{a--c})~\cite{Olstad2019CiliaryDevelopment}. We
first ask whether we can accurately represent these flow patterns with
computational fluid dynamics driven by cardiac pulsations and motile
cilia lining the surface of the brain ventricles.

We model the net forces induced by motile cilia as a steady
bidirectional traction acting on parts of the ventricular wall
(\Crefpanels{fig:fig1}{d}). Simulations show that this traction alone
leads to a partial compartmentalization of the ventricular system,
with large-scale vortex structures in the anterior, middle and
posterior ventricles (\Crefpanels{fig:fig1}{e}). The highest flow
speeds ($26.7$ \textmu m/s) occur in the vicinity of the dorsal cilia
in the middle ventricle. To model the cardiac pulsations, we impose a
pulsatile pressure difference between the anterior and posterior
boundaries (\Crefpanels{fig:fig1}{d, \fixme{g}}). The resulting flow
is pulsatile and directional (\Crefpanels{fig:fig1}{\fixme{f}}), and
in the absence of the cilia, the vortex structures vanish. The
velocity magnitude peaks at $4.8$ \textmu m/s. CSF initially flows
rostrocaudally before changing direction to caudorostral in the middle
of the cardiac cycle.

When combining the contributions, we recognize flow features
persisting throughout the cardiac cycle, now with vortex structures in
the anterior and middle ventricles and directional flow in the
posterior ventricle (\Crefpanels{fig:fig1}{h--j}, \fixme{Supplementary
  video 1}). These characteristic patterns suggest that the flow in
the anterior and middle ventricles is primarily driven by the motile
cilia, while the cardiac pressure pulsations dominate within the
posterior ventricle and in the ducts connecting the ventricular
compartments. The CSF flow speed peaks at the time of peak caudal
cardiac pulsatile flow, when both the tangential traction induced by
the cilia and the pulsatile pressure-driven flow are aligned with the
rostrocaudal axis, reaching up to 27.9 \textmu m/s dorsally in the
posterior part of the middle ventricle. The flow Reynolds number is
$\mathrm{Re} \approx 0.004$, justifying the assumption of Stokes
flow. In summary, the computational CSF flow model reproduces the flow
features observed in vivo~\cite{Olstad2019CiliaryDevelopment}.

@@

%% Over one cardiac cycle, the CSF flow speed reaches up to
%% 28.0 \textmu m/s, dorsally in the posterior part of the middle
%% ventricle. This peak is attained at the time instant coinciding with
%% the time of peak caudal cardiac pulsatile flow,

%% Using the observed maximum velocity magnitude and choosing the height
%% of the middle ventricle ($110$ \textmu m) as a characteristic length
%% scale yields a Reynolds number $\mathrm{Re} \approx 0.004$, justifying
%% the assumption of Stokes flow.

%We model the compartmentalized and pulsatile CSF flow observed in vivo~(\Crefpanels{fig:fig1}{b, c}) by the Stokes equations, and solve the equations numerically using a finite element method.

%To characterize CSF flow patterns predicted by
%computational fluid dynamics within the brain ventricles,

%we first consider the simulated CSF velocity and pressure when only
%modeling cilia motion, disregarding cardiac pressure pulsations.

@@

%We next explore the impact of these CSF flow patterns and magnitudes
%on solute transport within the brain ventricles.

Proteins, neurotransmitters and other molecular substances
intrinsically diffuse within the brain ventricles, but are also
advected by the flow of CSF. A natural question is which of these
mechanisms that dominate for any given type of solute. To address this
question, we simulate the distribution and evolution of three
different solutes injected within a region of interest (ROI 1) in the
dorsal region of the middle ventricle (\Crefpanels{fig:fig2}{a}),
subject to the CSF flow induced by motile cilia and cardiac pressure
pulsations. In particular, we consider three different molecular
diffusion coefficients, corresponding to a large extracellular vesicle ($D_1 = 2.17
\times 10^{-12}$ m$^2$/s), a 90 kDa protein STM-GFP ($D_2 = 5.75
\times 10^{-11}$ m$^2$/s), and a 26 kDa photoconvertible protein
Dendra2 ($D_3 = 1.15 \times 10^{-10}$ m$^2$/s).

%As expected, transport occurs more rapidly with increasing diffusion
%coefficient.

@@

%% To quantify the balance of advection versus diffusion as transport
%% mechanisms, we calculate the Péclet numbers $\mathrm{Pe}_1=519$,
%% $\mathrm{Pe}_2=14.7$ and $\mathrm{Pe}_3=7.4$ for $D_1$, $D_2$ and
%% $D_3$, respectively.  Since $\mathrm{Pe}>1$, this indicates that
%% advection is the dominant transport mechanism on a global scale.

@@

To further quantify the solute transport, we examine the mean
concentrations $\cbar_i(t)$ in the six ROIs over time
(\Crefpanels{fig:fig2}{e--g}) and, in addition, the times $\hat{t}_i$
when the mean concentrations first exceed a threshold value
$\hat{c}_i$ (\Crefpanels{fig:fig2}{h}).  Higher diffusion coefficients
result in more rapid transport (\Crefpanels{fig:fig2}{e--g}), with
lower times-to-threshold in ROIs 2--6 (\Crefpanels{fig:fig2}{h}). For
all three diffusion coefficients, the solute first spreads within the
middle ventricle, covering ROIs 2--4, before appearing in the anterior
ventricle (ROI 5) and finally reaching the posterior ventricle
(ROI 6). Notably, the extracellular vesicles, associated with the smallest diffusion
coefficient, do not reach the posterior ventricle (ROI 6) within the
simulated time frame~\Crefpanels{fig:fig2}{b, e}.

%% For all three diffusion coefficients, the dynamics of ROI 5
%% is significantly delayed compared to ROI 2--4,
%% and ROI 6 is the slowest.

%The dynamics at the site of photoconversion (ROI 1)
%is logarithmic and follows from imposing \Cref{eq:photoconversion_curve}.

%% For ROIs 2--6, we observe faster kinetics with higher diffusion coefficients
%% (\Crefpanels{fig:fig2}{e--g}),
%% which results in smaller times-to-threshold
%% (\Crefpanels{fig:fig2}{h}).

@@

Comparing the intensity changes with the simulated concentration
profiles in ROIs 1--6 (\Crefpanels{fig:fig3}{a, b}) show that the
model predictions agree well with the experimental measurements. We
note minor differences however. The simulated values of $\cbar_i$ at
the final time (0.80, 0.76, 0.86, 0.60 and 0.38 in ROIs 2--6, respectively)
are generally higher than the corresponding experimental $\Delta F$
values (0.63, 0.71, 0.67, 0.29 and 0.18). Moreover, we observe
differences in the distribution within the middle ventricle,
especially with regard to ROIs 2 and 3. Overall, we note that the time
to reach a 25\% threshold agree well between the model predictions and
experimental values (\Crefpanels{fig:fig3}{c, d}).

%% We obtained very similar kinetics curves (\Crefpanels{fig:fig3}{a, b})
%% and time-to-threshold values (\Crefpanels{fig:fig3}{c, d}) when
%% comparing our experiments (\Crefpanels{fig:fig3}{a, c}) with
%% simulation (\Crefpanels{fig:fig3}{b, d}).

%% Importantly, since the initial condition of the simulation is a mean
%% concentration of zero, the mean concentration $\cbar$ can be
%% interpreted as the change in mean concentration and is thus a similar
%% metric to $\Delta F$.

\begin{figure}%[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure3_compare_experiments_and_simulations.png}
    \caption{
    \textbf{a.} Relative change in fluorescence intensity $\Delta F(t)$ 
    in each region of interest (ROI)
    observed during experiments of photoconversion of Dendra2 proteins in ROI 1. 
    Bold lines are mean values of the cohort and shaded regions cover one standard devation errors
    in the mean.
    \textbf{b.} Mean concentration $\cbar(t)$ in each ROI in a
    simulation of Dendra2 protein transport with
    the diffusion coefficient $D_3$.
    \textbf{c.} Time to $\Delta F(t)$ reaches a threshold value of 0.25 in
    one representative zebrafish larva.
    Time to $\cbar(t)$ reaches a threshold value of 0.25 for
    a transport simulation with diffusion coefficient $D_3$.
    We have excluded the posterior-most regions of the geometries,
    because the values of $\Delta F$ and $\cbar$ never exceeded the threshold
    value of 0.25 in these regions.}
    \label{fig:fig3}
\end{figure}
% Final Delta F values
%0.6322650156762044 ROI 2
%0.7059449124175479 ROI 3
%0.6694500367160006 ROI 4
%0.28986700094045437 ROI 5
%0.18155055624229066 ROI 6
% Final mean c values ROIs 2-6: 0.7637806  0.72025859 0.81903216 0.56204164 0.34926411

@@

Ciliary motion is integral to physiological CSF flow,
and absence of cilia has been associated with defective
brain development and various pathologies~\cite{Eichele2020Cilia-drivenVentricle,
Olstad2019CiliaryDevelopment, Faubel2016Cilia-basedVentricles,
Guirao2010CouplingCilia, Afzelius2004CiliaRelatedDiseases,
Sawamoto2006NewBrain, Yoshiba2014RolesSymmetry,
Hirokawa2006NodalAsymmetry}.
As observed both experimentally and computationally,
the absence of motile cilia alters the CSF flow patterns. 
We next investigate the implications of these changes on intraventricular
transport.

We first experimentally measured the movement of photoconverted
Dendra2 signal in control and motile-cilia mutant (\emph{smh})
zebrafish. In ROIs 2 and 4, which measure Dendra2 distribution
downstream of the photoconversion site, we observed a slower increase
of fluorescence intensity (\Crefpanels{fig:fig4}{a}). The mean times-to-threshold
in the \emph{smh} mutants were $45\%$ (ROI 2) and $25\%$ (ROI 4) higher
compared to controls (\Crefpanels{fig:fig4}{b}). We found no significant differences
in the Dendra2 distribution in ROIs 3, 5, and 6 between
\emph{smh} mutants and controls.

@@

The CSF velocities are most strongly
affected when removing cilia in the dorsal part of the middle ventricle,
as the maximum velocity magnitude drops from 27.9 \textmu m/s to 7.9 \textmu m/s.
In vivo, these cilia generate the strongest flow at
this developmental stage of the fish~\cite{Olstad2019CiliaryDevelopment}.
In contrast, removing cilia in the anterior ventricle or the ventral region of the middle
ventricle has minimal impact, and results in maximum speeds
of 27.9 \textmu m/s and 27.4 \textmu m/s, respectively.
Of the total tangential forces exerted by the cilia, ventral cilia in the middle ventricle
contribute with 34\%, while the dorsal cilia exert 52\% of the total tangential
force~(\Crefpanels{fig:fig5}{b}). The magnitude of the force contributions are thus
more similar than the difference in their impact on the maximum velocity magnitudes.

@@

\subsection{Modification of ventricular geometry impacts solute distribution}
Variations in ventricular morphology impact CSF flow patterns~\cite{Olstad2019CiliaryDevelopment}.
To assess how these variations influence solute distribution,
we consider four alternative ventricular geometries:
one where the duct connecting the anterior and middle ventricles is constricted
(66 \% reduction in cross-sectional area);
one where the duct connecting the middle and posterior ventricles is constricted
(33 \% reduction);
one where the middle ventricle is shrunk in the lateral direction (43 \% reduction);
and the combination of the above three modifications (\Crefpanels{fig:fig6}{a, b}). In these
four modified geometries the volume is reduced by 3.9\%, 4.1\%, 1.4\% and 9.4\%, respectively,
compared to the original geometry.

Constricting the connection between the anterior and middle ventricle reduces
solute transport to the anterior ventricle (ROI 5),
and the mean concentration in ROIs 2--4 and 6 slightly increases, compared to the original geometry
(\Crefpanels{fig:fig6}{c}). This is notwithstanding a
9.5\% reduction in ventral cilia force magnitude
(\Crefpanels{fig:fig6}{d}), which are the cilia that propel CSF and solute towards ROI 6.
Constricting the mid-hind brain connection
leads to less transport into the posterior ventricle (ROI 6),
and slightly more transport towards the anterior ventricle (ROI 5).
Faster transport to the anterior ventricle (ROI 5) and slower transport to the
posterior ventricle (ROI 6) is also the case for the geometry with a
shrunk middle ventricle.
Here the transport to the posterior ventricle (ROI 6) is reduced,
in spite of a 6.6\% increase in the ventral cilia forces.
Additionally, for the shrunk middle configuration, the transport upstream of CSF flow towards the
ventral-posterior region of the middle ventricle (ROI 3) is slowed down.

>>>>

\author[1,*]{Halvor Herlyng}
\author[1]{Ada J.~Ellingsrud}
\author[1]{Miroslav Kuchta} 
\author[2]{Inyoung Jeong}
\author[1,3]{Marie E.~Rognes}
\author[2,*]{Nathalie Jurisch-Yaksi}
\affil[1]{Department of Numerical Analysis and Scientific Computing, Simula Research Laboratory, Oslo, Norway} % Kristian Augusts gate 23, 0164
\affil[2]{Department of Clinical and Molecular Medicine, Norwegian University of Science and Technology, Trondheim, Norway} % Erling Skjalgsons gate 1, 7491
\affil[3]{K.~G.~Jebsen Center for Brain Fluid Research, Oslo, Norway}
\affil[*]{\href{mailto:hherlyng@simula.no}{hherlyng@simula.no}, \href{mailto:nathalie.jurisch-yaksi@ntnu.no}{nathalie.jurisch-yaksi@ntnu.no}}
\date{\vspace{-2em}}

@@

% Abstract should be written in the present tense and impersonal style (i.e., avoid we), and be at most 200 words long
\begin{abstract}
Cerebrospinal fluid (CSF) is integral to brain function, as the fluid
provides mechanical support for the brain and the flow of CSF helps
distribute nutrients, neurotransmitters and metabolites throughout
the central nervous system. CSF flow is partly driven
by beating of motile cilia, hair-like structures that line the ependymal cell layer
of the walls of the brain ventricles. Despite the physiological importance of CSF flow,
the underlying mechanisms of cilia-mediated flow and transport
in the brain ventricles remain yet to be comprehensively resolved.
In this study, we analyze and evaluate the role of motile cilia
in intraventricular CSF flow and transport. We use the geometry
of the zebrafish larval brain ventricles, for which we have detailed knowledge of cilia properties and CSF motion velocities
and frequencies, to develop finite element methods that model flow and transport.
The computational model is validated by in vivo experiments that monitor transport
of a photoconvertible protein secreted in the brain ventricles.
Our results show that while cilia contribute to advection of large particles,
diffusion plays a significant role in the transport of small molecules such as proteins.
We also found that cilia location and the geometry of the ventricular system impact solute distribution.
Altogether, our work presents a computational framework that can be applied to other
ventricular systems, together with new concepts of how molecules are transported within the brain and its ventricles.

@@

% I2: What do we know about the problem and the topic?
The flow of CSF in the ventricular system is driven by multiple
physiological processes, including via the beating motion of motile
cilia, CSF secretion by the choroid plexus, and pulsatile pressure gradients generated by the
cardiac cycle, the respiration, neural activity, and bodily
movements~\cite{Brodal2016, del2010ependymal, Ringers2020Role, Vinje2019RespiratoryMeasurements,
Kurtcuoglu2005ComputationalSystem, Olstad2019CiliaryDevelopment,
macaulay2022cerebrospinal, mestre2018flow}.
Cilia are hair-like structures found in a vast diversity of
biological organisms and in several human organs,
of which the brain is one of them~\cite{Fulford1986Muco-ciliaryLung,
Pacherres2022CiliaryProduction, Jahn1972LocomotionProtozoa,
Brennen1977FluidFlagella, Ringers2020Role, spassky2013motile,
DGama2025MotileBrain, mitchell2007evolution,
Tsukita2012CoordinatedFeet, BLAKE1974MechanicsMotion,
Reiten2017Motile-Cilia-MediatedComputations, Thouvenin2020OriginCanal}. 
The ventricular surface of the brain, known as the ependyma, is lined by
motile cilia that beat in a coordinated pattern at frequencies of
10--40 Hz to mediate fluid flow~\cite{mitchell2007evolution,
Ringers2020Role, spassky2013motile, Ringers2023NovelEpithelia, roth2025structure}. 
Given their microscopic size, ranging from
5 to 15 microns, cilia contribute primarily to flow near the
ependyma in the human brain~\cite{Siyahhan2014FlowVentricles, Olstad2019CiliaryDevelopment,
Ringers2020Role, Faubel2016Cilia-basedVentricles}. In contrast, cardiac,
respiratory and delta waves at lower frequencies induce pulsatile flow
at the larger scales in the brain ventricular system, subarachnoid space, and
subarachnoid perivascular
spaces~\cite{Vinje2019RespiratoryMeasurements, eide2024functional,
Causemann2025, fultz2019coupled, mestre2018flow, daversin2020mechanisms}.

@@

% I4: How do we address the problem and what do we find out?
In this study, we combined computational and experimental approaches to investigate
the role of motile cilia in intraventricular CSF flow and transport.
We model the flow and transport by partial differential equations and solve
them numerically with finite element methods to compute flow patterns and
solute concentration dynamics. A geometry representing the zebrafish larval brain ventricles
was used, for which we have detailed knowledge of cilia properties and CSF motion velocities
and frequencies~\cite{Olstad2019CiliaryDevelopment}. 
Notably, this early developmental stage allows us to simplify our model
by omitting CSF secretion, which starts at later stages~\cite{Jeong2024TheZebrafish},
while still being well-conserved with mammals~\cite{DGama2025MotileBrain,
Olstad2019CiliaryDevelopment, Ringers2020Role, DGama2021Diversity, jurisch2020radial}. 
The computational model is validated with experimental data by monitoring in vivo
the dynamics of a photoconvertible fluorescent protein Dendra2
expressed within the brain ventricles.
Our results show that diffusion plays a major role in the transport of small solutes,
whereas cilia are paramount for the movement of larger particles.
We also found that the location of cilia, as well as changes to the ventricular geometry,
largely impact solute distribution.
In summary, our work presents a computational
framework that can be applied to other
ventricular systems in the future, together with new concepts
of how cilia distribute molecules and
particles within the brain ventricles.

@@

In vivo particle tracking reveals characteristic CSF flow patterns in
zebrafish larval brain ventricles
(\Crefpanels{fig:fig1}{a--c})~\cite{Olstad2019CiliaryDevelopment}.
We attempt to accurately represent these flow patterns with
computational fluid dynamics driven by cardiac pulsations and motile
cilia lining the surface of the brain ventricles.

We assume that viscous forces are dominating the intraventricular CSF flow,
and model the flow by the Stokes equations.
The net forces imparted by motile cilia on the CSF are modeled as a steady
bidirectional traction acting on parts of the ventricular wall
(\Crefpanels{fig:fig1}{d}), and we assume free velocity slip at the walls.
Simulations show that this traction alone
leads to a partial compartmentalization of the ventricular system,
with large-scale vortex structures in the anterior, middle and
posterior ventricles (\Crefpanels{fig:fig1}{e}). The highest flow
speeds ($26.7$ \textmu m/s) occur in the vicinity of the dorsal cilia
in the middle ventricle. To model the cardiac pulsations, we impose a
pulsatile pressure difference between the anterior and posterior
boundaries (\Crefpanels{fig:fig1}{d, \fixme{g}}). The resulting flow
is pulsatile and directional (\Crefpanels{fig:fig1}{\fixme{f}}), and
in the absence of the cilia, the vortex structures vanish. The
velocity magnitude peaks at $4.8$ \textmu m/s. CSF initially flows
rostrocaudally before changing direction to caudorostral in the middle
of the cardiac cycle.

When combining cilia motion and cardiac pulsations, we recognize flow features
persisting throughout the cardiac cycle, now with vortex structures in
the anterior and middle ventricles and directional flow in the
posterior ventricle (\Crefpanels{fig:fig1}{h--j}, \fixme{Supplementary
  video 1}). These characteristic patterns suggest that the flow in
the anterior and middle ventricles is primarily driven by the motile
cilia, while the cardiac pressure pulsations dominate within the
posterior ventricle and in the ducts connecting the ventricular
compartments. The CSF flow speed peaks at the time of peak caudal
cardiac pulsatile flow, when both the tangential traction induced by
the cilia and the pulsatile pressure-driven flow are aligned with the
rostrocaudal axis, reaching up to 27.9 \textmu m/s dorsally in the
posterior part of the middle ventricle. The flow Reynolds number is
$\mathrm{Re} \approx 0.004$, justifying the assumption of Stokes
flow. In summary, the computational CSF flow model reproduces the flow
features observed in vivo~\cite{Olstad2019CiliaryDevelopment}.

@@

@@

Proteins, neurotransmitters and other molecular substances
intrinsically diffuse within the brain ventricles, but are also
advected by the flow of CSF. A natural question is which of these
mechanisms that dominate for any given type of solute. To address this
question, we simulate the distribution and evolution of three
different diffusing solutes injected within a region of interest (ROI 1) in the
dorsal region of the middle ventricle (\Crefpanels{fig:fig2}{a}),
subject to the CSF flow induced by motile cilia and cardiac pressure
pulsations. In particular, we consider three different molecular
diffusion coefficients, corresponding to a large extracellular vesicle ($D_1 = 2.17
\times 10^{-12}$ m$^2$/s), a 90 kDa protein STM-GFP ($D_2 = 5.75
\times 10^{-11}$ m$^2$/s), and a 26 kDa photoconvertible protein
Dendra2 ($D_3 = 1.15 \times 10^{-10}$ m$^2$/s).

@@

To further quantify the solute transport, we examine the mean
concentrations $\cbar_i(t)$ in six ROIs over time
(\Crefpanels{fig:fig2}{e--g}) and, in addition, the times $\hat{t}_i$
when the mean concentrations first exceed a threshold value
$\hat{c}_i$ (\Crefpanels{fig:fig2}{h}).  Higher diffusion coefficients
result in more rapid transport (\Crefpanels{fig:fig2}{e--g}), with
lower times-to-threshold in ROIs 2--6 (\Crefpanels{fig:fig2}{h}). For
all three diffusion coefficients, the solute first spreads within the
middle ventricle, covering ROIs 2--4, before appearing in the anterior
ventricle (ROI 5) and finally reaching the posterior ventricle
(ROI 6). Notably, the extracellular vesicles, associated with the smallest diffusion
coefficient, do not reach the posterior ventricle (ROI 6) within the
simulated time frame~\Crefpanels{fig:fig2}{b, e}.

@@

@@

Comparison of the intensity changes with the simulated concentration
profiles in ROIs 1--6 (\Crefpanels{fig:fig3}{a, b}) shows that the
model predictions agree well with the experimental measurements. We
note minor differences however. The simulated values of $\cbar_i$ at
the final time (0.80, 0.76, 0.86, 0.60 and 0.38 in ROIs 2--6, respectively)
are generally higher than the corresponding experimental $\Delta F$
values (0.63, 0.71, 0.67, 0.29 and 0.18). Moreover, we observe
differences in the distribution within the middle ventricle,
especially with regard to ROIs 2 and 3. Overall, we note that the time
to reach a 25\% threshold agree well between the model predictions and
experimental values (\Crefpanels{fig:fig3}{c, d}).

\begin{figure}%[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure3_compare_experiments_and_simulations.png}
    \caption{
    \textbf{a.} Relative change in fluorescence intensity $\Delta F(t)$ 
    in each region of interest (ROI)
    observed during experiments of photoconversion of Dendra2 proteins in ROI 1. 
    Bold lines are mean values of the cohort and shaded regions cover one standard devation errors
    in the mean.
    \textbf{b.} Mean concentration $\cbar(t)$ in each ROI in a
    simulation of Dendra2 protein transport with
    the diffusion coefficient $D_3$.
    \textbf{c.} Time to $\Delta F(t)$ reaches a threshold value of 0.25 in
    one representative zebrafish larva.
    Time to $\cbar(t)$ reaches a threshold value of 0.25 for
    a transport simulation with diffusion coefficient $D_3$.
    We have excluded the posterior-most regions of the geometries,
    because the values of $\Delta F$ and $\cbar$ never exceeded the threshold
    value of 0.25 in these regions.}
    \label{fig:fig3}
\end{figure}
% Final Delta F values
%0.6322650156762044 ROI 2
%0.7059449124175479 ROI 3
%0.6694500367160006 ROI 4
%0.28986700094045437 ROI 5
%0.18155055624229066 ROI 6
% Final mean c values ROIs 2-6: 0.80, 0.76, 0.86, 0.60 and 0.38

@@

Ciliary motion is integral to physiological CSF flow,
and absence of cilia has been associated with defective
brain development and various pathologies~\cite{Eichele2020Cilia-drivenVentricle,
Olstad2019CiliaryDevelopment, Faubel2016Cilia-basedVentricles,
Guirao2010CouplingCilia, Afzelius2004CiliaRelatedDiseases,
Sawamoto2006NewBrain, Yoshiba2014RolesSymmetry,
Hirokawa2006NodalAsymmetry}.
As observed both experimentally and computationally,
the absence of motile cilia alters CSF flow patterns. 
We next investigate the implications of these changes on intraventricular
transport.

We experimentally measured the movement of proteins by monitoring
photoconverted Dendra2 signal in control zebrafish and 
motile-cilia mutant \emph{schmalhans} (\emph{smh}) zebrafish.
In ROIs 2 and 4, which quantify Dendra2 distribution
downstream of the photoconversion site, we observed a slower increase
of fluorescence intensity (\Crefpanels{fig:fig4}{a}). The mean times-to-threshold
in the \emph{smh} mutants were $45\%$ (ROI 2) and $25\%$ (ROI 4) higher
compared to controls (\Crefpanels{fig:fig4}{b}). We found no significant differences
between \emph{smh} mutants and controls
in the Dendra2 distribution in ROIs 3, 5, and 6.

@@

The CSF velocities are most strongly
affected when removing cilia in the dorsal part of the middle ventricle,
as the maximum velocity magnitude drops from 27.9 \textmu m/s to 7.9 \textmu m/s.
In vivo, these dorsal cilia generate the strongest flow at
this developmental stage of the fish~(\Crefpanels{fig:fig1}{c})~\cite{Olstad2019CiliaryDevelopment}.
In contrast, removing cilia in the anterior ventricle or the ventral region of the middle
ventricle has minimal impact, and results in maximum speeds
of 27.9 \textmu m/s and 27.4 \textmu m/s, respectively.
Of the total tangential forces exerted by the cilia, ventral cilia in the middle ventricle
contribute with 34\%, while the dorsal cilia exert 52\% of the total tangential
force~(\Crefpanels{fig:fig5}{b}). The magnitude of these force contributions are thus
more similar than the difference in their impact on the maximum velocity magnitudes.

@@

\subsection{Modification of ventricular geometry impacts solute distribution}
Variations in ventricular morphology impact CSF flow patterns~\cite{Olstad2019CiliaryDevelopment}.
To assess how these variations influence solute distribution,
we consider four alternative ventricular geometries:
one where the duct connecting the anterior and middle ventricles is constricted
(66 \% reduction in cross-sectional area);
one where the duct connecting the middle and posterior ventricles is constricted
(33 \% reduction);
one where the middle ventricle is shrunk in the lateral direction (43 \% reduction);
and the combination of the above three modifications (\Crefpanels{fig:fig6}{a, b}).
In these four modified geometries, the total volume is reduced by
3.9\%, 4.1\%, 1.4\% and 9.4\%, respectively,
compared to the original geometry.

Constricting the connection between the anterior and middle ventricle reduces
solute transport to the anterior ventricle (ROI 5),
and the mean concentration in ROIs 2--4 and 6 slightly increases, compared to the original geometry
(\Crefpanels{fig:fig6}{c}). This is notwithstanding a
9.5\% reduction in ventral cilia force magnitude
(\Crefpanels{fig:fig6}{d}), which are the cilia that propel CSF and solute towards ROI 6.
Constricting the mid-hind brain connection
leads to less transport into the posterior ventricle (ROI 6),
and slightly more transport towards the anterior ventricle (ROI 5).
Faster transport to the anterior ventricle (ROI 5) and slower transport to the
posterior ventricle (ROI 6) is also the case for the geometry with a
shrunk middle ventricle.
This reduction in the transport to the posterior ventricle (ROI 6)
happens in spite of a 6.6\% increase in the ventral cilia forces.
Additionally, for the shrunk middle configuration, the transport upstream of CSF flow towards the
ventral-posterior region of the middle ventricle (ROI 3) is slowed down.
